# Preamble

## Dependencies

```{r}
#| label: load-libs
#| message: false
#| warning: false
#| results: hide
source("utils.R")
library(spatstat)
library(seg)
library(stats)
library(RANN)
library(ggplot2)
library(spatstat.geom)
```

## Setup

```{r}
#| label: load-data
spe <- readRDS("../data/spe.rds")
```

# Load data

```{r}
sub <- spe[, spe$sample_id == 0.01]
(pp <- .ppp(sub, marks = "cluster_id"))
```


# Visualize 

Plot using spatstat

```{r, fig.width=12, fig.height=12}
plot(pp)
```

Discrete labels, called "marks", can be extracted with `marks`.

```{r}
marks(pp)[1:5]
```

Marks can be removed from a point pattern with `unmark`.

```{r}
#| fig-width: 10
#| fig-height: 10
plot(unmark(pp))
```


The marks are now set to a factor.

```{r}
marks(pp) <- factor(marks(pp))
```

Using `split` a point pattern can be directly split by mark type.

```{r}
#| fig-width: 10
#| fig-height: 10
ppls <- split(pp)
plot(ppls)
```


And with `superimpose` point patterns can be combined. 

```{r}
do.call(superimpose,ppls)
```

The summary of `ppp` (point pattern) object returns general properties, plus intensities, combined and per mark type.

```{r}
summary(pp)
```

If the point pattern has discrete marks, the function `intensity` will return the intensity for each type. 
```{r}
intensity(pp)
```

To get the overall intensity the individual intensities can be summed up.

Assuming that the the multi-type process is first order stationary (i.e. each sub-process is stationary) the individual intensities sum up to the intensity of the unmarked point process.

```{r}
sum(intensity(pp))
intensity(unmark(pp))
```

However, the stationarity assumption in our case in inappropriate (c.f. Ependymal cells)

*TODO* What does this imply and should we test for it?

The kernel density estimates can again be showed for all points together or per type.

```{r}
#| fig-width: 10
#| fig-height: 10
plot(density(pp))
```

```{r, fig.width=10 ,fig.height=10}
plot(density(ppls))
plot(density(split(pp), se=TRUE)$SE)
```

This indicates that our dataset seems to be inhomogeneous, i.e. the different cell types have different intensity functions

The relative risk, i.e. the probability of observing a given celltype at a given location, can be calculated with `relrisk`. The bandwidth for smoothing is calculated with `bw.relrisk` and might need to be adjusted.


```{r}
#| fig-width: 10
#| fig-height: 10
rp <- relrisk(pp,se=TRUE)
plot(rp$estimate)
plot(rp$SE)
```


```{r, eval=FALSE, echo=FALSE}
im_sum <- im.apply(rp$estimate,sum)
im_sum <- signif(im_sum,9)
all(im_sum==1)
```

Using the `relrisk` function we can get the dominant mark for different regions of the tissue of interest.

```{r, eval=FALSE, echo=FALSE}
#| fig-width: 10
#| fig-height: 10
rpd <- relrisk(pp,diggle = TRUE)
dom <- im.apply(rpd,which.max)
dom <- eval.im(factor(dom,levels=seq_along(levels(unique(marks(pp)))),labels = levels(unique(marks(pp)))))
plot(dom,las=2,main="Dominant mark")
```
Under the assumption that the points were generated by a Poisson process we can identify hotspots and coldspots by comparing the standard error of the `relrisk` function to the theoretical null distribution of points.

[Source](https://idblr.rbind.io/post/pvalues-spatial-segregation/)

```{r}
#Code source: https://idblr.rbind.io/post/pvalues-spatial-segregation/

f1 <- rp
# Significant p-values assumming normality of the Poisson process
## relrisk() computes standard errors based on asymptotic theory, assuming a Poisson process
  alpha <- 0.05                           # alpha
  z <- qnorm(alpha/2, lower.tail = F)     # z-statistic
  f1$u <- f1$estimate + z*f1$SE           # Upper CIs
  f1$l <- f1$estimate - z*f1$SE           # Lower CIs
  mu_0 <- as.vector(table(spatstat.geom::marks(pp))/pp$n) # null expectations by type
  f1$p <- f1$estimate # copy structure of pixels, replace values
  for (i in 1:length(f1$p)) {
    f1$p[[i]]$v <- factor(ifelse(mu_0[i] > f1$u[[i]]$v, "lower",
                                 ifelse( mu_0[i] < f1$l[[i]]$v, "higher", "none")),
                          levels = c("lower", "none", "higher"))
  }

# Plot significant p-values
plot(f1$p, main = "Significant difference\n to Poisson process alpha = 0.05")
```

# Random labeling

A test of segregation of marks can be performed with `segregation.test`. It is based on the assumption of random labeling and the computed test statistic measures the similarity of the distribution of marks. The randomization is done by permuting the marks of the points while keeping the locations constant.

```{r}
#| warning: false
pp

segregation.test(pp,nsim=19)
```

In this case there is clear indication that the labeling of marks is not random.

# Measures of spatial segregation

The package `seg` implements different measures of spatial segregation. Spatial segregation measures encompass two distinct concepts. The first describes the evenness versus clustering which is independent of the overall composition of a sample and describes the distribution of individuals. The second describes isolation versus exposure which depends on the overall composition of a sample and describes the degree to which different groups are isolated or exposed to one another.

Spatial segregation measures developped by Reardon and O'Sullivan (2004) [Reardon, S. F. and O'Sullivan, D. (2004) Measures of spatial segregation. Sociological Methodology, 34, 121-162.]. D, R and H measure spatial evenness while P measures the spatial exposure.

- Dissimilarity index (D) describes how different the composition of a local environment is, on average, from the composition of the total population. It ranges from 0 to 1, where 0 indicates perfect integration, and 1 indicates complete segregation.
- Relative diversity (R) is a measure of how much less diverse each individuals local environments are, on average, compared to the total population. It ranges from 0 to 1, where 0 indicates perfect integration, and 1 indicates complete segregation.
- Information theory index (H) is a measure of how much less diverse individuals local environments are, on average, than is the total population. It can be interpreted as a measure of the variation in the diversity of each individuals local environment. It ranges from 0 to 1, where 0 indicates no segregation, and 1 indicates maximum possible segregation, i.e. all subjects in each local environment belong to one group only.
- The spatial exposure index (P) expresses the degree of spatial exposure of group m to group n. It is the average percentage of group n in the local environments of each member of group m and ranges from 0 (no exposure) to 1 (maximal exposure).

Index of Spatial Proximity developed by White (1983) [White, M. J. (1983). The measurement of spatial segregation. The American Journal of Sociology, 88, 1008-1018.].

- This numerical value indicates the degree of segregation. A value of 1 indicates evenness in the sample, and values greater than 1.0 indicate clustering. If the index value is smaller than one, it indicates an unusual form of segregation (i.e., some groups are closer to other groups). It compares the average distance between members of one group with that between all individuals, irrespective of group assignment. It may change depend on the definition of distance.


```{r}
pp_sel <- pp

# create an indicator matrix
model.matrix <- matrix(0, nrow=length(marks(pp_sel)), ncol=length(unique(marks(pp_sel))))
model.matrix[cbind(seq_along(marks(pp_sel)), marks(pp_sel))] <- 1
colnames(model.matrix) <- unique(marks(pp_sel))

spseg(pp_sel,
      data = model.matrix,
      smoothing = 'kernel')

# Index of spatial proximity
paste0('Index of spatial proximity: ',
isp(pp_sel,
    data = model.matrix))
```
### Example of clear spatial segregation

```{r}
selection <- c('Astrocyte', 'Ependymal')
pp_sel <-  subset(pp, marks %in% selection, drop = TRUE)


plot(pp_sel)

# create an indicator matrix
model.matrix <- matrix(0, nrow=length(marks(pp_sel)), ncol=length(unique(marks(pp_sel))))
model.matrix[cbind(seq_along(marks(pp_sel)), marks(pp_sel))] <- 1
colnames(model.matrix) <- unique(marks(pp_sel))

spseg(pp_sel,
      data = model.matrix,
      smoothing = 'kernel')

# Index of spatial proximity
paste0('Index of spatial proximity: ',
       isp(pp_sel,
           data = model.matrix))
```

### Example of less segregation


```{r}
selection <- c('Astrocyte', 'Microglia')
pp_sel <-  subset(pp, marks %in% selection, drop = TRUE)


plot(pp_sel)
plot(unmark(pp_sel))

# create an indicator matrix
model.matrix <- matrix(0, nrow=length(marks(pp_sel)), ncol=length(unique(marks(pp_sel))))
model.matrix[cbind(seq_along(marks(pp_sel)), marks(pp_sel))] <- 1
colnames(model.matrix) <- unique(marks(pp_sel))

spseg(pp_sel,
      data = model.matrix,
      smoothing = 'kernel')

# Index of spatial proximity
paste0('Index of spatial proximity: ',
       isp(pp_sel,
           data = model.matrix))
```

# Correlation and spacing

## Distances and nearest neighbors

Investigating the nearest neighbor distance between point for all combinations of marks can be done as follows:

```{r}
d <- nndist(pp,by=marks(pp))
a <- aggregate(d,by=list(from=marks(pp)),min)
a
```

Annother option is to either use the `K` nearest neighbors or count the number of points of a given type with a distance closer than `R`. Both can be done with the function `marktable`, where the argument `R` stands for the neighborhood radius and `N` for number of neighbors.

```{r}
mt <- marktable(pp,N=5,collapse = TRUE)
signif(mt/rowSums(mt),2)

mt <- marktable(pp,R=10,collapse = TRUE)
signif(mt/rowSums(mt),2)

```

Additionally the function `markstat` can be used to calculate more complicated expressions than just the sum.

## Nearest neighbor correlations

A overall correlation between marks can be calculated with `nncorr`. It returns two values: `unnormalised`, which is the probability that a point and its nearest neighbor have the same type and `normalised`, which divides the unnormalised probability by the probability of random labeling. So a value close to 1 indicates random labeling. A value much larger than 1 means neighboring point are often of the same type. 

```{r}
nncorr(pp)
```

The R package `spdep` (https://cran.r-project.org/package=spdep) contains many functions to calculate statistics based on nearest neighbors.

Another possibility is to work with nearest neighborhood contingency tables to do statistical tests using the `dixon` function from the R package dixon. It allows to calculate the statistic "segregation of species" S which indicates either random labeling (if S=0), attraction (if S<0) or seggregation (if S>0).
```{r}
if(require(dixon)){
  out <- dixon::dixon(as.data.frame(pp))
  out$tablaZ
}
```


### Cuzick-Edwards' k-NN test

```{r}
# Function to calculate Cuzick and Edwards' T_k statistic for a marked ppp
calculate_Tk_marked <- function(ppp, k_range) {
  # Check if the input point pattern is a marked ppp object
  if (!is.ppp(ppp) || is.null(marks(ppp))) {
    stop("The point pattern must be a marked ppp object.")
  }
  
  # Get the unique marks
  unique_marks <- unique(marks(ppp))
  
  # Initialize a data frame to store the T_k statistic for each mark and k
  results <- data.frame()
  
  # Calculate the T_k statistic for each mark and k
  for (mark in unique_marks) {
    # Subset the point pattern by the current mark
    ppp_marked <- subset(ppp, marks == mark)
    
    # Separate the cases and controls
    cases <- subset(ppp, marks == mark)
    controls <- subset(ppp, marks != mark)
    # Remark cases and control
    marks(cases) <- as.factor('case')
    marks(controls) <-  as.factor('control')
    
    # Combine the cases and controls into a single point pattern
    combined_ppp <- superimpose(cases, controls)
    
    # Loop through all k's
    for (k in k_range) {
      
      # Index of cases
      idx <- marks(combined_ppp) == 'case'
      
      # Sum occurrences of cases
      T_k <- sum(marktable(combined_ppp, N=k)
                 [marks(combined_ppp) == 'case','case'])
      
      # NOT IN ORGINAL FORMULA: divide by total number of points
      T_k <- T_k / sum(marks(combined_ppp) == 'case')
      
      # Add the results to the data frame
      results <- rbind(results, 
                       data.frame(Mark = as.character(mark),
                                  K = k, Tk = T_k))
    }
  }
  
  # Plot the T_k statistic vs. k for each mark
  ggplot(results, aes(x = K, y = Tk, color = Mark)) +
    geom_point() +
    labs(title = "",
         x = "k",
         y = "Cuzick and Edwards' T_k statistic",
         color = "Mark")
}

# Apply the function to a marked ppp
calculate_Tk_marked(pp_sel, c(1:10))

plot(split(pp_sel))
```

### Scan statistic and friends


## Summary functions for pairs of types

```{r}
# for a faster runtime we will use a subset of the data
selection <- c('Astrocyte', 'OD Mature', 'Ependymal', 'Microglia', 'Pericytes')
pp_sel <-  subset(pp, marks %in% selection, drop = TRUE)

```



Similar to the simple case without marks it is possible do estimate summary functions. More specifically the summary functions between different marks can be calculated. 

These summary functions assume that the multitype process is stationary which might not be an appropriate assumption in spatial omcis data.

### Cross K-function

```{r}
Kall <- alltypes(rescale(pp_sel), Kcross)
plot(Kall)
```

The diagonal shows the K-function for the different marks, with indication of Poisson or non-Poisson point processes. Off-diagonal panels give indication of independence of points when the number of points follows the expected K-function and does not imply that the points are Poisson


### Cross L-function:

Similiar to the the K-function, the cross L-function can be defined for all pairs of points.

```{r}
#| fig-width: 10
#| fig-height: 10
lc <- alltypes(pp_sel, Lcross)
plot(lc, .-r~r)
```

```{r}
#| fig-width: 10
#| fig-height: 10
lce <- alltypes(pp_sel, Lcross, envelope = TRUE, nsim=9)
plot(lce, .-r~r)
```
- TODO: not sure what the off-diagonal plots show us

### Cross pair-correlation function

```{r}
#| fig-width: 10
#| fig-height: 10
pcfc <- alltypes(pp_sel, pcfcross)
plot(pcfc)
```


### Mark connection function

The mark connection function is essentially the cross pair-correlation function divided by the unmarked pair-correlation function. It can be interpreted as the conditional probability that two points a distance r apart have labels of type 1 and of type 2, given the presence of those points. 

```{r}
#| fig-width: 10
#| fig-height: 10

mc <- alltypes(pp_sel, markconnect)
plot(mc)
```

The dashed lines indicate expected values under random labeling, revealing that nearby points are more likely to have different types than expected by chance. This positive association between different cell types does not necessarily imply dependence, as it could be influenced by a negative association between cells of the same type.

### Cross F-function (empty space function)

The cross F-function is the cumulative distribution function of the distance from a location to the nearest point of the same type. Assumes that the process is stationary.

```{r}
#| fig-width: 5
#| fig-height: 10
fc <- alltypes(pp_sel, Fest)
plot(fc)
```

### Cross G-function (Nearest-neighbor function)

The cross G-function is the cumulative distribution function of the distance from a location to the nearest point of another type. If the points are independent of each other the G and F function are identical. Assumes that the process is stationary.

```{r}
#| fig-width: 10
#| fig-height: 10
gc <- alltypes(pp_sel, Gcross)
plot(gc)
```

### Cross J-function 

The value of 1 is consistent with CSR and independence between different points.

```{r}
#| fig-width: 10
#| fig-height: 10
jc <- alltypes(pp_sel, Jcross)
plot(jc)
```

### Dot functions

For each K-, G- and J- function there also exist dot functions which is measuring distances from points of one type to points of any type.

```{r}
jd <- alltypes(pp_sel, Jdot)
plot(jd)
```


## Summary function within and between types

### Mark equality function

The Mark or Type Equality function for a stationary multitype point process measures the correlation between types of two points separated by distance r. If k < 1, points at distance r are less likely than expected to be of the same type. If > 1, they are more likely to be of the same type.

```{r}
#| fig-width: 10
#| fig-height: 10
me <- markcorr(pp_sel)
plot(me)
```

### I-function 

```{r}
#| fig-width: 10
#| fig-height: 10
ic <- Iest(pp_sel)
plot(ic)
```

## Summary functions for non stationary processes

The cross and dot functions described above assume a underlying stationary point process. If this assumption is not given, the interpretation of the result might be incorrect.

The inhomogeneous equivalents of the described functions require the different intensities as inputs in `spatstat`. If they are not given the intensities are estimated using kernel smoothing.  



### Inhomogeneous L-function
```{r}
#| fig-width: 10
#| fig-height: 10
lci <- alltypes(pp_sel, Lcross.inhom)
plot(lci, .-r~r)
```

# Tests of randomness and independence

# Multitype models and processes

## Multitype Poisson models

Similar to the unmarked case marked point patterns can be fitted to specific modeling assumptions. The simples case is the homogeneous multitype poisson process. It can be fit again with `ppm` but this time using "marks" as the covariate, which essentially means a different intensity is assumed for each type. 

```{r}
ppfit <- ppm(pp~marks)
ppfit
```

To fit a nonstationary multitype poisson process a additional covariates including the special variables "x" and "y" can be used. For example:

```{r}
ppfit <- ppm(pp~marks * polynom(x,y,1))
ppfit
```

- TODO

## Multitype Gibbs models

## Multitype Cox processes

## Multitype Cluster processes

# Correlation stationarity

```{r}
selection <- c('OD Mature')
pp_sel <-  subset(pp, marks %in% selection, drop = TRUE)


odrho <- rhohat(unmark(pp_sel), "x", method="tr")
odlambda <- predict(odrho)

od4 <- quantess(unmark(pp_sel), "x", 2)
od42 <- nestsplit(pp_sel, od4, ny=3)

plot(od42)

od42$inten <- factor(as.integer(od42$f1) <= 1, labels=c("Hi","Lo"))

studpermu.test(od42, pts ~ inten, summaryfunction=Kscaled, minpoints = 10)

studpermu.test(od42, pts~ inten, summaryfunction=Kinhom,
               lambda=odlambda, minpoints = 10)
```

```{r}
selection <- c('Ependymal')
pp_sel <-  subset(pp, marks %in% selection, drop = TRUE)


odrho <- rhohat(unmark(pp_sel), "x", method="tr")
odlambda <- predict(odrho)

od4 <- quantess(unmark(pp_sel), "x", 2)
od42 <- nestsplit(pp_sel, od4, ny=3)

plot(od42)

od42$inten <- factor(as.integer(od42$f1) <= 1, labels=c("Hi","Lo"))

studpermu.test(od42, pts ~ inten, summaryfunction=Kscaled, minpoints = 10)

studpermu.test(od42, pts~ inten, summaryfunction=Kinhom,
               lambda=odlambda, minpoints = 10)
```

# Appendix

## Session info

```{r}
#| label: session-info
sessionInfo()
```
