# Lattice data analysis -- univariate methods for imaging-based data

In this vignette we will show:

- Univariate lattice data analysis methods for imaging-based approaches.

- This includes global metrics on the entire field of view and local variants thereof. 

- The use case is a CosMx data set from @heHighplexImagingRNA2022.

- The `R` implementations rely on the `r BiocStyle::Biocpkg('Voyager')` package. The data is represented as `r BiocStyle::Biocpkg('SpatialFeatureExperiment')` [@mosesVoyagerExploratorySinglecell2023]. Complementary resources using this data and methods are found in the [Voyager CosMx, vignette](https://pachterlab.github.io/voyager/articles/vig4_cosmx.html).

- `Python` implementations rely on the the packages [esda](https://pysal.org/esda/), [pysal](https://pysal.org/) and [squidpy](https://squidpy.readthedocs.io/en/stable/) [@reyPySALPythonLibrary2010; @pallaSquidpyScalableFramework2022]. Data representation rely on the [anndata](https://anndata.readthedocs.io/en/stable/) structure [@virshup2024anndata].

## Dependencies

::: {.panel-tabset}

### `R`

```{r}
#| label: load-libs
#| message: false
#| warning: false
#| results: hide
source("utils.R")

roma_colors <- data.frame(roma_colors = scico::scico(256, palette = 'roma'))
write.csv(roma_colors , "../misc/roma_colors.csv")

theme_set(theme_minimal())
```

### `Python`

```{python}
import numpy as np
import scanpy as sc
import squidpy as sq
from esda.moran import Moran, Moran_Local
from esda.geary import Geary
from esda import Geary_Local, G_Local, LOSH
from esda.getisord import G
from scipy.stats import false_discovery_control
from libpysal.weights import W, KNN
from libpysal.cg import KDTree
import pandas as pd
import matplotlib.pyplot as plt
from matplotlib.colors import ListedColormap, LinearSegmentedColormap
import warnings

warnings.filterwarnings("ignore")

plt.rcParams['font.family'] = 'Futura'
df_cmap_continuous = pd.read_csv("../misc/roma_colors.csv", index_col=0)
cmap_continuous = LinearSegmentedColormap.from_list("roma", list(df_cmap_continuous["roma_colors"])).reversed()
```

::: 

Until now, we have considered the cells to be represented in a point pattern. However, as cells have a shape and area, this might be an oversimplification in some cases. Alternatively, we can rely on the segmentation of individual cells that are available for various datasets. The outline of each cell is represented by a polygon and the collection of all cells can be seen as an irregular lattice. Unlike a regular lattice (e.g., spot-based spatial transcriptomics data), the sample areas in an irregular lattice can have different sizes and are not necessarily regularly distributed over the sample space.

For this representation of the cells we will rely on the `r BiocStyle::Biocpkg('SpatialFeatureExperiment')` package. For preprocessing of the dataset and code examples we refer the reader to the [vignette](https://pachterlab.github.io/voyager/articles/vig4_cosmx.html) of the `r BiocStyle::Biocpkg('voyager')` package [@mosesVoyagerExploratorySinglecell2023]. The `r BiocStyle::Biocpkg('voyager')` package provides wrapper functions around the package `r BiocStyle::CRANpkg('spdep')` [@pebesmaSpatialDataScience2023] that work directly on the `r BiocStyle::Biocpkg('SpatialFeatureExperiment')` object.

## Setup and Preprocessing

::: {.panel-tabset}

### `R`

```{r}
#taken from https://pachterlab.github.io/voyager/articles/vig4_cosmx.html
(sfe <- HeNSCLCData())

# Empty cells
colData(sfe)$is_empty <- colData(sfe)$nCounts < 1
# Select, sum negative control probes
(neg_inds <- str_detect(rownames(sfe), "^NegPrb")) %>% sum


colData(sfe)$prop_neg <- colSums(counts(sfe)[neg_inds,])/colData(sfe)$nCounts
# Remove low quality cells
sfe <- sfe[,!sfe$is_empty & sfe$prop_neg < 0.1]
# Re-calculate stats
rowData(sfe)$is_neg <- neg_inds
# log Counts
sfe <- logNormCounts(sfe)

# save for python
colData(sfe)$CenterX_global_px <- spatialCoords(sfe)[,1]
colData(sfe)$CenterY_global_px <- spatialCoords(sfe)[,2]
```

For some examples we will show a subset of the tissue.

```{r}
bbox_use <- st_as_sfc(st_bbox(c(xmin = 3200, xmax = 16800, ymin = 155200, ymax = 166200)))
```

```{r}
sfe <- sfe[,st_intersects(colGeometries(sfe)$centroids, bbox_use, sparse = FALSE)]
ann <-  zellkonverter::SCE2AnnData(sfe, X_name = "counts")
anndata::write_h5ad(ann, "../data/imaging_sfe.h5ad")
```

### `Python`

```{python}
adata = sc.read_h5ad("../data/imaging_sfe.h5ad")
adata.obsm["spatial"] = np.column_stack([adata.obs["CenterX_global_px"], adata.obs["CenterY_global_px"]])
# normalise counts
adata.raw = adata.copy()
sc.pp.normalize_total(adata)
sc.pp.log1p(adata)
# invert the spatial coordinates
adata.obsm["spatial"][:, 1] = adata.obsm["spatial"][:, 1].max() - adata.obsm["spatial"][:, 1]
adata
```

:::

Here we set the arguments for the examples below. We will continue with the two features *KRT17* (basal cells) and *TAGLN* (smooth muscle cells) and specify a $k$-nearest neighbour weight matrix with $k=5$.

::: {.panel-tabset}

### `R`

```{r}
features <- c("KRT17", "TAGLN")
colGraphName <- "knn5"
colGeometryName <- "centroids"
pointsize <- 1
```

### `Python`

```{python}
# predefine genes
features = ["KRT17", "TAGLN"]
figsize = (len(features)*10, 10)
pointsize = 30
```

:::

::: {.panel-tabset}

### `R`

```{r}
#| fig-width: 15
#| fig-height: 20
plotSpatialFeature(sfe, features,
                   colGeometryName = "centroids", 
                   nrow = 2, size = 2, scattermore = FALSE) +
  theme_void()
```

### `Python`

```{python}
#| fig-width: 15
#| fig-height: 20
fig, axes = plt.subplots(len(features), 1, figsize=(10, 15), layout = "tight")
for feature, ax in zip(features, axes):
    sq.pl.spatial_scatter(adata, color=feature, shape=None, size=10, cmap="Blues", ax=ax, fig=fig, use_raw=False)
    ax.set_axis_off()
```

:::

{{< include theory/03-theory-irreg-lattice.qmd >}}

{{< include theory/04-theory-univar-lattice.qmd >}}

# Appendix

## Session info

```{r}
#| label: session-info
sessionInfo()
```
