# Lattice data analysis -- multivariate methods for imaging-based data

In this vignette we will show:

- Multivariate lattice data analysis methods for imaging-based approaches.

- This includes global metrics on the entire field of view and local variants thereof. 

- The use case is a CosMx data set from @heHighplexImagingRNA2022.

- Complementary resources using this data and methods are found in the [Voyager CosMx vignette](https://pachterlab.github.io/voyager/articles/vig4_cosmx.html), [Voyager bivariate vignette](https://pachterlab.github.io/voyager/articles/bivariate.html) and [Voyager multivariate vignette](https://pachterlab.github.io/voyager/articles/localc.html).

## Dependencies

```{r}
#| label: load-libs
#| message: false
#| warning: false
#| results: hide
source("utils.R")
theme_set(theme_minimal())
```

For this representation of cells, we will rely on the `SpatialFeatureExperiment` package. For preprocessing of the dataset, we refer the reader to the [vignette](https://pachterlab.github.io/voyager/articles/vig4_cosmx.html) of the `voyager` package.


```{r}
#taken from https://pachterlab.github.io/voyager/articles/vig4_cosmx.html
(sfe <- HeNSCLCData())

# Empty cells
colData(sfe)$is_empty <- colData(sfe)$nCounts < 1
# Select, sum negative control probes
(neg_inds <- str_detect(rownames(sfe), "^NegPrb")) %>% sum


colData(sfe)$prop_neg <- colSums(counts(sfe)[neg_inds,])/colData(sfe)$nCounts
# Remove low quality cells
sfe <- sfe[,!sfe$is_empty & sfe$prop_neg < 0.1]
# Re-calculate stats
rowData(sfe)$is_neg <- neg_inds
# log Counts
sfe <- logNormCounts(sfe)
```

```{r}
#| include: false
#| eval: true
bbox_use <- st_as_sfc(st_bbox(c(xmin = 3200, xmax = 16800, ymin = 155200, ymax = 166200)))
sfe <- sfe[,st_intersects(colGeometries(sfe)$centroids, bbox_use, sparse = FALSE)]
```


<!-- [MR: do these need to be recalculated?] -->

<!-- Markers: KRT17 (basal cells), C1QA (macrophages), IL7R (T cells), TAGLN (Smooth muscle cells) -->

In this vignette we are highlighting lattice data analysis approaches for multivariate observations. We will show the metrics related to a ligand-receptor pair, CEACAM6 and EGFR which was identified in the original publication of the CosMx dataset [@heHighplexImagingRNA2022].


```{r}
#| fig-width: 15
#| fig-height: 10
plotSpatialFeature(sfe, c("CEACAM6"),
                   colGeometryName = "centroids", 
                   ncol = 2, size = 2, scattermore = FALSE) + 
  theme_void()
```


```{r}
#| fig-width: 15
#| fig-height: 10
plotSpatialFeature(sfe, c("EGFR"),
                   colGeometryName = "centroids", 
                   ncol = 2, size = 2, scattermore = FALSE) + 
  theme_void()
```

{{< include theory/03-theory-irreg-lattice.qmd >}}

Here we set the arguments for the examples below.

```{r}
features <- c("KRT17", "TAGLN")
colGraphName <- "knn5"
colGeometryName <- "centroids"
segmentation <- "cellSeg"
```

With a defined spatial weight matrix, one can calculate multivariate spatial metrics. We will consider both global and local bivariate observations as well as local multivariate spatial metrics.

{{< include theory/05-theory-multivar-lattice.qmd >}}

## Session info

```{r}
#| label: session-info
sessionInfo()
```
