# Lattice data analysis -- multivariate methods for imaging-based data

In this vignette we will show:

- Multivariate lattice data analysis methods for imaging-based approaches.

- This includes global metrics on the entire field of view and local variants thereof. 

- The use case is a CosMx data set from @heHighplexImagingRNA2022.

- The `R` implementations rely on the `r BiocStyle::Biocpkg('Voyager')` package. The data is represented as `r BiocStyle::Biocpkg('SpatialFeatureExperiment')` [@mosesVoyagerExploratorySinglecell2023]. Complementary resources using this data and methods are found in the [Voyager CosMx vignette](https://pachterlab.github.io/voyager/articles/vig4_cosmx.html), [Voyager bivariate vignette](https://pachterlab.github.io/voyager/articles/bivariate.html) and [Voyager multivariate vignette](https://pachterlab.github.io/voyager/articles/localc.html).

- `Python` implementations rely on the the packages [esda](https://pysal.org/esda/), [pysal](https://pysal.org/) and [squidpy](https://squidpy.readthedocs.io/en/stable/) [@reyPySALPythonLibrary2010; @pallaSquidpyScalableFramework2022]. Data representation rely on the [anndata](https://anndata.readthedocs.io/en/stable/) structure [@virshup2024anndata].

## Dependencies

::: {.panel-tabset}

### `R`

```{r}
#| label: load-libs
#| message: false
#| warning: false
#| results: hide
source("utils.R")

roma_colors <- data.frame(roma_colors = scico::scico(256, palette = 'roma'))
write.csv(roma_colors , "../misc/roma_colors.csv")

theme_set(theme_light())
```

### `Python`

```{python}
import numpy as np
import scanpy as sc
import squidpy as sq
from esda.moran import Moran_BV, Moran_Local_BV
from esda.lee import Spatial_Pearson, Spatial_Pearson_Local
from esda.geary_local_mv import Geary_Local_MV
from scipy.stats import false_discovery_control
from libpysal.cg import KDTree
from libpysal.weights import W, KNN
import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
from matplotlib.colors import ListedColormap, LinearSegmentedColormap
import warnings

warnings.filterwarnings("ignore")

plt.rcParams['font.family'] = 'Futura'
df_cmap_continuous = pd.read_csv("../misc/roma_colors.csv", index_col=0)
cmap_continuous = LinearSegmentedColormap.from_list("roma", list(df_cmap_continuous["roma_colors"])).reversed()
```

For this representation of cells, we will rely on the `SpatialFeatureExperiment` package. For preprocessing of the dataset, we refer the reader to the [vignette](https://pachterlab.github.io/voyager/articles/vig4_cosmx.html) of the `voyager` package.

:::

## Setup and Preprocessing

::: {.panel-tabset}

### `R`


```{r}
#taken from https://pachterlab.github.io/voyager/articles/vig4_cosmx.html
(sfe <- HeNSCLCData())

# Empty cells
colData(sfe)$is_empty <- colData(sfe)$nCounts < 1
# Select, sum negative control probes
(neg_inds <- str_detect(rownames(sfe), "^NegPrb")) %>% sum


colData(sfe)$prop_neg <- colSums(counts(sfe)[neg_inds,])/colData(sfe)$nCounts
# Remove low quality cells
sfe <- sfe[,!sfe$is_empty & sfe$prop_neg < 0.1]
# Re-calculate stats
rowData(sfe)$is_neg <- neg_inds
# log Counts
sfe <- logNormCounts(sfe)

# save for python
colData(sfe)$CenterX_global_px <- spatialCoords(sfe)[,1]
colData(sfe)$CenterY_global_px <- spatialCoords(sfe)[,2]
```

```{r}
#| include: false
#| eval: true
bbox_use <- st_as_sfc(st_bbox(c(xmin = 3200, xmax = 16800, ymin = 155200, ymax = 166200)))
sfe <- sfe[,st_intersects(colGeometries(sfe)$centroids, bbox_use, sparse = FALSE)]
```

```{r}
sfe <- sfe[,st_intersects(colGeometries(sfe)$centroids, bbox_use, sparse = FALSE)]
ann <-  zellkonverter::SCE2AnnData(sfe, X_name = "counts")
anndata::write_h5ad(ann, "../data/imaging_sfe.h5ad")
```

### `Python`

```{python}
adata = sc.read_h5ad("../data/imaging_sfe.h5ad")
adata.obsm["spatial"] = np.column_stack([adata.obs["CenterX_global_px"], adata.obs["CenterY_global_px"]])
# normalise counts
adata.raw = adata.copy()
sc.pp.normalize_total(adata)
sc.pp.log1p(adata)
# invert the spatial coordinates
adata.obsm["spatial"][:, 1] = adata.obsm["spatial"][:, 1].max() - adata.obsm["spatial"][:, 1]
adata
```


::: 

<!-- [MR: do these need to be recalculated?] -->

<!-- Markers: KRT17 (basal cells), C1QA (macrophages), IL7R (T cells), TAGLN (Smooth muscle cells) -->

In this vignette we are highlighting lattice data analysis approaches for multivariate observations. We will show the metrics related to *KRT17* (basal cells) and *TAGLN* (Smooth muscle cells) [@heHighplexImagingRNA2022].


```{r}
#| fig-width: 15
#| fig-height: 10
plotSpatialFeature(sfe, c("KRT17"),
                   colGeometryName = "centroids", 
                   ncol = 2, size = 2, scattermore = FALSE) + 
  theme_void()
```


```{r}
#| fig-width: 15
#| fig-height: 10
plotSpatialFeature(sfe, c("TAGLN"),
                   colGeometryName = "centroids", 
                   ncol = 2, size = 2, scattermore = FALSE) + 
  theme_void()
```

Here we set the arguments for the examples below.

::: {.panel-tabset}

### `R`

```{r}
features <- c("KRT17", "TAGLN")
colGraphName <- "knn5"
colGeometryName <- "centroids"
segmentation <- "cellSeg"
plotsize = 1.5
```

### `Python`

```{python}
# predefine genes
features = ["KRT17", "TAGLN"]
figsize = (10, 7)
pointsize = 12
```

:::

{{< include theory/03-theory-irreg-lattice.qmd >}}

With a defined spatial weight matrix, one can calculate multivariate spatial metrics. We will consider both global and local bivariate observations as well as local multivariate spatial metrics.

{{< include theory/05-theory-multivar-lattice.qmd >}}

# Appendix

## Session info

```{r}
#| label: session-info
sessionInfo()
```
